---
title: "Pre-Analysis Plan: Ideological Biases of the Crowd?"
date: "`r format(Sys.time(), '%d %B, %Y - %X (%Z)')`"
output: 
  pdf_document:
    toc: yes
    number_sections: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
urlcolor: blue
linkcolor: blue
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{amsmath}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage[utf8]{inputenc}
---
```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
#library(printr)
library(tidyverse)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
```

\newpage

# Expectations

**Ideological Bias hypothesis** (_H1a_): The larger the ideological distance between respondent and the party, the less likely respondents code stances according to the party’s position.

**Ideological Bias hypothesis** (_H1b_): The effect of H1a is stronger for the underspecified condition.

**Ideological Overinterpretation hypothesis** (_H2a_):  The larger the ideological distance between respondent and the party, the more likely respondents interpret underspecified sentences as stance.

**Political Knowledge Overinterpretation hypothesis** (_H2b_): The more political knowledge, the more likely people interpret underspecified sentences as stance. 

**Masking Solution hypothesis** (_H3a_):  Masking reduces the effect of respondents’ ideological position for coding stances according to the party’s position.

**Masking Solution hypothesis** (_H3b_): Masking reduces the effect of respondents’ level of political knowledge for coding stances according to the party’s position. 

# Research Desing and Protocol

## Sample
We  will conduct this survey experiment in the Netherlands in April 2022.
The sample, recruited through [KiesKompas](https://www.kieskompas.nl/en/), will consist of 2,000 participants (based on the power analysis presented in Figure \ref{fig:power1}) of 18 years and older.
Kieskompas works with non-random opt-in respondents.
Therefore, I measure many demographic background variables (see [Section 3.2](#control-variables)).
Balance checks will be conducted to demonstrate whether certain categories are over represented in a certain experimental group.
The study has been approved by the [Research Ethics Review Committee](https://fsw.vu.nl/nl/onderzoek/research-ethics-review/index.aspx) of the _Vrije Universiteit Amsterdam_ (see the approval [here](https://github.com/MarikenvdVelden/bias-experiment/blob/master/docs/2022-3-30-59.pdf)).
To ensure good quality of our data, one attention check (discussed in more detail in [Section 3.3](#attention-checks)) is included.

## Experimental Protocol
The study is conducted online and in Dutch.
Participants are told that they are taking part in a survey to get an overview of how Dutch people form their views on politics.
After reading an informed consent message participants are forwarded to the main questionnaire (or the survey will be terminated if they do not agree to the consent form).

First, participants complete a set of pre-treatment variables (i.e. vote recall, issue positions of the issues used in the experiment (Climate, EU, Immigration, and Tax), ideology, and political knowledge).
This block ends with the attention check included in this survey.
When participants fail this attention check, a warning appears asking them to read the question again
carefully  and  to  answer  again.
Therefter, participants see the experimental stimuli.
The stimuli in the experiment are to show respondents fully specified or underspecified sentences on the issue position of a political actor or a masked political actor (see Section Treatment for more details). 

## Power Analysis
As detailed in [Section 4](#analysis), we conduct an OLS regression for the four sentences.
The manipulation, i.e. the masking condition vs. the political actor, as well as the specification of the sentence (i.e. _fully specified_ or _underspecified_) as well as respondents ideological distance to the party and their levels of political knowledge are the main independent variables.
  
To calculate power for the hypotheses, the R package `DeclareDesign` is used [@blair2019declaring].
The effect sizes are between $b = 0.2$ and $b = 0.1$ -- i.e. a small effect visualized by the purple and blue lines in Figure \ref{fig:power1}.
The hypothesis are directional, Figure \ref{fig:power1} therefore displays one-tailed tests with $\alpha = 0.05$.
The power analysis shows that testing hypotheses 1 and 2 requires a sample size of $2,000$ participants, who all annotate the four sentences, i.e. $8,000$ participants in total, (x-axis) to reach very high levels of power, i.e. $>95\%$, (indicated by the black dashed line in Figure \ref{fig:power}). 
  
```{r power, fig.align = 'center', out.width = "100%", fig.cap = "\\label{fig:power1}Power Analysis", fig.topcaption=TRUE}
source(here::here("docs/poweranalysis.R"))
p
```
  
# Measures

## Dependent Variables
We rely on whether or not a party's (implied) stance is coded according to the party’s position (H1 and H3) as well as whether or not the statement is coded as a stance at all (H3). For each issue, we ask the respondent `what is according to the sentence above the position of [ACTOR]?`, with the answer categories: `in favor`, `against`, `no stance`, `don't know`.

## Treatment
Respondents are randomly assigned to either view a political party as an actor, or a masked condition, where they see `X` as an actor; simultaneously, respondents see either a fully specified sentence or a underspecified sentence, in which one needs additional information to interpret the positon on an actor.
Table \ref{tab:conditions} gives an overview of the variations in treatment  in the survey as well as their English translations.

```{r conditions, topcaption=TRUE}
d <- tibble(Condition = c(rep("Specified", 4),
                          rep("Underspecified", 4)), 
            `Wording ENG` = c("[PVV/X] says immigration should be made harder.",
                              "[GreenLeft/X] says nitrogen emissions need to be reduced.",
                              "[Labour Party/X] says tax rate should go up for highest earners.",
                              "[Forum for Democracy/X] says that membership in the European Union has been especially bad for the Netherlands so far.".
                              "[PVV/X] says many immigrants are coming this way.",
                              "[GreenLeft/X] says nitrogen policy must be different.",
                              "[Labour Party/X] says tax system must be changed.",
                              "[Forum for Democracy/X] says the Netherlands should have a different role in the European Union."),
            `Wording NL` = c("[PVV/X] zegt dat immigratie moeilijker gemaakt moet worden.",
                             "[GroenLinks/X] zegt dat stikstofuitstoot meer tegengegaan moet worden.",
                             "[PvdA/X] zegt dat het belastingtarief voor de hoogste inkomens omhoog moet.",
                             "[Forum voor Democratie/X] zegt dat het lidmaatschap van de Europese Unie tot nu toe vooral slecht geweest voor Nederland is.",
                             "[PVV/X] zegt dat veel immigranten deze kant op komen.",
                             "[GroenLinks/X] zegt dat het stikstofbeleid anders moet.",
                             "[PvdA/X] zegt dat het belastingstelsel moet worden aangepast.",
                             "[Forum voor Democratie/X] zegt dat Nederland een andere rol in de Europese Unie moet hebben."))

kbl(d, booktabs =T, caption = "\\label{tab:conditions}Survey Questions - Experimental Conditions") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = F, fixed_thead = T, position = "center") %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "7cm") %>%
  column_spec(3, width = "7cm")
```

## Control Variables

As control variables, the following _demographics_ are measured post-treatment: gender, age, education, geographical region, level of urbanness, employment, and income.
For the analysis, only variables that are unbalanced over the experimental conditions will be included.
Table \ref{tab:demographics} gives an overview of the questions asked in the survey as well as their English translations.

- _Gender_ is measured as `sex`. The answer categories are `Male` (value of `1`), `Female` (value of `0`), and `No answer` (value of `999`).

- _Age_ is measured using 6 categories: `17 or younger`, `18--29`, `30--39`, `40--49`, `50--59`, `60--74`. 

- _Education_ is measured as the highest successfully completed level of education, recoded into four categories: `low`, `middle`, `high`, and `none`.
I create dummy variables for each level of education with the lowest category as base category.

- _Employment_ Respondents were asked which category of employment -- `Full-time employed`, `Part-time employed`, `Entrepreneur`, `Unemployed and searching for a job`, `Unemployed and not searching for a job or incapacitated`, `Housewife/Househusband or else`, `Retired`, `Student or full-time education` -- applied most to them. 

- _Income_ Respondents were questioned on their monthly income in bins of €500 -- `€500 or less`, `€501-€1000`, `€1001-€1500`, `€1501-€2000`, `€2001-€2500`, `€2501-€3000`, `€3001-€3500`, `€3501-€4000`, `€4501-€7500`, `€7501 or more` -- as well as giving them the options of `won't  say` and `don't know`.

- _Geographical region_ is measured using the _Nielsen districs_, dividing the Netherlands into 1) the 3 major cities plus suburbs, Amsterdam (plus Diemen, Ouder-Amstel, Landsmeer, Amstelveen), Rotterdam (plus Schiedam, Capelle aan den IJssel, Krimpen aan den IJssel, Nederlek, Ridderkerk, Barendrecht, Albrandswaard) and The Hague (plus Leidschendam, Voorburg, Rijswijk, Wassenaar, Wateringen); 2) West (Noord-Holland, Zuid-Holland and Utrecht (excluding the major cities and their suburbs; 3) North (Groningen, Friesland and Drenthe), 4) East (Overijssel, Gelderland and Flevoland); and South (Zeeland, Noord-Brabant and Limburg).


```{r demographics, topcaption=TRUE}
d <- tibble(Variable = c("gender (D1)", "age (D2)", 
                         "region (D3)", "job (D5)", "income (D6)",
                         "education (D7)"), 
            `Wording ENG` = c("What is your gender?", "What is your year of birth?",
                             "In which region do you live?",
                             "What applies most to you? Are you...",
                              "Can you indicate which income groups your net monthly household income falls into?",
                             "What is your highest level of education?", ),
            `Wording NL` = c("Wat is uw geslacht?", "Wat is uw geboortejaar?", 
                             "In welke regio woont u?",
                             "Wat is het meest op u van toepassing? Bent u...",
                             "Kunt u aangeven in welke inkomensgroepen uw netto maandelijks huishoudinkomen valt?", 
                             "Wat is uw hoogst GENOTEN opleiding",))

kbl(d, booktabs =T, caption = "\\label{tab:demographics}Survey Questions - Demographics") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = F, fixed_thead = T, position = "center") %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "7cm") %>%
  column_spec(3, width = "7cm")
```

In addition, pre-treatment, respondents' ideological position, political knowledge, vote recall, and position on the issues migration, climate, tax, and EU are measured (see Table \ref{tab:pret}).
Those variables will only be included in the analyses if balance checks indicate they are necessary.
Moreover, the variables will be used to explore heterogeneous relationships. 

- _Ideological position_ is measured using an 11-point scale ranging from left (`0`) to right (`10`).

- _Vote Recall_ Respondents were asked which party they voted for in the 2021 parliamentary elections. The options were 1) all parties that were elected into parliament -- `Bij1`, `BoerBurgerBeweging`, `CDA`, `ChristenUnie`, `D66`, `Denk`, `Forum for Democracy`, `JA21`, `GroenLinks`, `PvdA`, `Animal Rights Party`, `PVV`, `SGP`, `SP`, `VOLT`,`VVD`, `50Plus Party` -- 2) another party; 3) blanco vote; and 4) a `Don't know` option.


```{r pret, topcaption=TRUE}
d <- tibble(Variable = c("vote recall (PT1)", "introduction",
                         "migration (PT2)", "climate (PT3)",
                         "tax (PT4)", "EU (PT5)", "RILE (PT6)",
                         "political-knowledge1 (PT7_1)", 
                         "political-knowledge2 (PT7_2)", 
                         "political-knowledge3 (PT7_3)", 
                         "political-knowledge4 (PT7_4)", 
                         "political-knowledge5 (PT7_5)",
                         "political-knowledge5 (PT7_6)"), 
            `Wording ENG` = c( "Which party did you vote for during the last parliamentary elections of March 2021?",
                               "To what extent do you agree with the following statements:",
                               "There are too many immigrants in the Netherlands.",
                               "The climate crisis is exaggerated.",
                               "The tax rate for the highest earners should go up",
                               ""
                               "In politics there is often talk about left and right. Where would you put yourself on this scale?",
                               "Which parties currently form the coalition government?",
                               "Which of the following is the current Finance Minister in the Rutte cabinet?",
                               "Which political party was the second-largest party in the last Lower House elections on March 17?",
                               "By whom are members of the Senate elected?",
                               "Who makes or is making new laws in the Netherlands?",
                               "If the government and the House of Representatives pass a law that is not in line with the Dutch Constitution, may the courts disapprove it?"),
            `Wording NL` = c("Op welke politieke partij heeft u bij de vorige Tweede Kamerverkiezingen van maart 2021 gestemd?",
                             "In hoeverre bent u het eens met onderstaande stellingen:",
                             "Er zijn te veel immigranten in Nederland.",
                             "De klimaatcrisis wordt overdreven.",
                             "Het belastingtarief voor de hoogste inkomens moet omhoog",
                             "Het lidmaatschap van de Europese Unie is tot nu toe vooral slecht geweest voor Nederland",
                             "In de politiek wordt vaak gesproken over links en rechts. Waar zou u uzelf op deze schaal plaatsen?",
                             "Welke partijen vormen op dit moment de coalitieregering?",
                            "Wie van de volgende personen is de huidige minister van Financiën in het kabinet Rutte?",
                            "Welke politieke partij was de op één na grootste partij tijdens de afgelopen Tweede Kamerverkiezingen van 17 maart?",
                            "Door wie worden de leden van de Eerste Kamer gekozen?",
                            "Wie maakt of maken in Nederland nieuwe wetten?",
                            "Als de regering en de Tweede Kamer een wet aannemen die niet overeenkomt met de Nederlandse Grondwet, mag de rechter deze wet dan afkeuren?"))

kbl(d, booktabs =T, caption = "\\label{tab:pret}Survey Questions - PreTreatment Questions") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = F, fixed_thead = T, position = "center") %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "7cm") %>%
  column_spec(3, width = "7cm")
```

- _Political knowledge_ is measured with six items from the DPES.

- _Position on migration_ is measured by asking people whether or not there are too many immigrants in the Netherlands using a 5-point Likert-scale (`fully disagree`, `disagree`, `neutral`, `agree`, `fully agree`).

- _Position on climate_ is measured by asking people whether or not the climate crisis is exaggerated using a 5-point Likert-scale (`fully disagree`, `disagree`, `neutral`, `agree`, `fully agree`).

- _Position on tax_ is measured by asking people whether or not the  tax rate for the highest earners should go up using a 5-point Likert-scale (`fully disagree`, `disagree`, `neutral`, `agree`, `fully agree`).


- _Position on the EU_ is measured by asking people whether or not membership in the European Union has been especially bad for the Netherlands so far using a 5-point Likert-scale (`fully disagree`, `disagree`, `neutral`, `agree`, `fully agree`).


## Attention Check
We include one attention checks in the survey.
This is asked just before respondents enter the round of the experimental treatments.
The attention checks are taken from @berinsky2014separating and adapted to the German context by the authors.
If a respondents fails the first attention check, a warning appears and the respondent can only continue with the survey once the respondent has correctly answered the question correctly.

**Attention Check**  When a big news story breaks people often go online to get up-to-the-minute details on what is going on.
We want to know which websites people trust to get this information.
We also want to know if people are paying attention to the question.
To show that you have read this much, please ignore the question and select Volkskrant and Metro as your two answers.
When there is a big news story, which is the one news website you would visit first? (Please
only choose one).
Eight (Dutch) news outlets are provided to choose from.
Respondents pass the attention check if they select _de Volkskrant_ and _Metro_.


## Exclusion Criteria
Participants are required to respond to each question.


# Analysis
We test the hypotheses formulated in [Section 1](#expectations) by fitting linear multivariate regressions pooling the issues.
In each model, we will estimate the coefficient for whether the actor in the sentence is masked or not, as well as for whether the sentence is fully specified or underspecified. We will add ideological distance from the party mentioned as well as political knowledge as interactions with the experimental conditions.
We will only add control variables in the analyses that are unbalanced, as explained in [Section 4.1](#balance-checks).

## Balance Checks
We will conduct a balance test based on demographics (age, gender, education, geographical region, level of urbanness,employment, and income), vote choice in the 2021 parliamentary elections, ideological self-placement, political knowledge, and positions on the issues, using the `cobalt` R package [@greifer2021]. 
If the groups are unbalanced on one of these variables -- i.e. standardized mean differences $<0.05$ -- we will add the covariates to the analyses.
We will use the code below to conduct the balance tests (see [here](https://github.com/MarikenvdVelden/compromise-punish/tree/main/src/analysis/balance-test.R) for the R script).

```{r balance, eval=FALSE, echo = TRUE}
covs <- d %>%
  mutate(treatment = sentence) %>%
  select(sentence, D1:D7, PT1:PT&) 

balanced <-bal.tab(treatment ~ factor(D1) + D2 +
                     factor(D3) + D4 + D5 + 
                     factor(D6) + D7 + PT1 + 
                     PT2 + PT3 + PT4 + PT5 + 
                     PT6 + PT7,
                     data = covs,
          thresholds = c(m = 0.05))[[1]] 
```

## Hypothesis 1
To test the _Ideological Bias hypothesis_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level.\

$$\begin{equation}\label{eq:pooled}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}ideological-distance-to-party_{r, i, t} +
    \beta_{4}political-knowledge_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}$$

We consider evidence for H1a when the effect of $\beta_{3}$ is postive and statistically significant.
If the Average Marginal Effect of $\beta_{3}$ is positive and statistically siginificant when $\beta_{2} = 1$ (i.e. condition is underspecidied), the second _Ideological Bias hypothesis_ (H1b) will be confirmed.




```{r h1, eval = FALSE, echo = TRUE}
h1a <- lmer(stance_correct ~ masked + specification + 
              ideological_distance_to_party + political_knowledge +
              (1 | issue), data= d)

h1b <- lmer(stance_correct ~ masked + 
              specification * ideological_distance_to_party + 
              political_knowledge + (1 | issue), data= d)
```

## Hypothesis 2
To test the _Overinterpretation Hypotheses_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level

$$\begin{equation}\label{eq:pooled}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}heterogeneous_treatment{r, i, t} +
    \beta_{4} specification * heterogeneous_treatment_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}$$

When the Average Marginal Effect of $\beta_{3}$ ideological distance is positive and statistically siginificant when $\beta_{2} = 1$ (i.e. condition is underspecidied), the _Ideological Overinterpretation hypothesis_ (H2a) will be confirmed.
When the Average Marginal Effect of $\beta_{3}$ political knowledge is positive and statistically siginificant when $\beta_{2} = 1$ (i.e. condition is underspecidied), the _Political Knowledge Overinterpretation hypothesis_ (H2b) will be confirmed.

```{r h2, eval = FALSE, echo = TRUE}
h2a <- lmer(stance ~ masked + specification * ideological_distance_to_party +
              political_knowledge + (1 | issue), data= d)

h2b <- lmer(stance ~ masked + 
              specification * ypolitical_knowledge + 
              ideological_distance_to_party + (1 | issue), data= d)
```

## Hypothesis 3
To test the _Masking Solution hypotheses_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level

$$\begin{equation}\label{eq:pooled}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}heterogeneous_treatment{r, i, t} +
    \beta_{4} masked * heterogeneous_treatment_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}$$

When the Average Marginal Effect of $\beta_{3}$ ideological distance is negatiive and statistically siginificant when $\beta_{2} = 1$ (i.e. condition is masked), the first _Masking Solution hypothesis_ (H3a) will be confirmed.
When the Average Marginal Effect of $\beta_{3}$ political knowledge is negative and statistically siginificant when $\beta_{2} = 1$ (i.e. condition is masked), the second _Masking Solution hypothesis_ (H3b) will be confirmed.

```{r h3, eval = FALSE, echo = TRUE}
h3a <- lmer(stance ~ masked + specification * ideological_distance_to_party +
              political_knowledge + (1 | issue), data= d)

h3b <- lmer(stance ~ masked + 
              specification * ypolitical_knowledge + 
              ideological_distance_to_party + (1 | issue), data= d)
```


## Statistical Significance
All the hypotheses are directional, and therefore all of the tests will be one-tailed.
We will use an $\alpha$-value of $0.05$ as the value for statistical significance in all models above.

  
\newpage
# References